\documentclass{article}

\usepackage{a4wide}
\usepackage{url}

\begin{document}

\author{Wojciech Mostowski\\
Radboud University Nijmegen\\
Department of Computing Science\\
Nijmegen, the Netherlands\\
e-mail: \url{woj@cs.ru.nl}}

\title{\textbf{The APDU Interface of the PKI Applet}}

\maketitle


\section{Introduction}

This document briefly describes the features and the APDU interface of
the Java Card PKI applet. The applet has been developed according to
the ISO7816 specification, for information not included here please
refer to~\cite{ISO4,ISO8,ISO15}, or contact the author. For this
applet a certain small selection of options from the ISO standard has
been implemented, as explained below.  The source code of the applet
and the host library is available from the SourceForge SVN repository
at:
\begin{center}
\url{https://javacardsign.svn.sourceforge.net/svnroot/javacardsign}
\end{center}
The SourceForge address of the PKI project is
\url{http://javacardsign.sourceforge.net}.  The current code in the
project has been developed mainly by Wojciech Mostowski,
\url{woj@cs.ru.nl}.

\section{Applet Specifications}

The current version of the applet and the host library implements the
following features:
\begin{itemize}
\item An ISO7816 file system for storing PKI files according to the
  Part 15 of the ISO7816 specification~\cite{ISO15}: private key
  directory, certificate directory, CA and user certificates, etc.  It
  is up to the personalisation software what files will be stored in
  the applet.  The applet support hierarchical file system including
  relative to current file selection or selection by path. Reading of
  each file can be user PIN protected.
\item PIN and PUC user authentication: a 4--20 characters long PIN
  code, and a 16 characters long PUC code. The PUC code lets the user
  to unblock a forgotten PIN code.
\item The applet does not support any kind of secure messaging for
  APDU communication.
\item The applet only communicates on the contact interface of the
  card.
\item Three different cryptographic operations: signing, decryption,
  and authentication.  Currently, the supported key type and length is
  RSA 1024 bit.\footnote{This is an ``artificial'' limitation.  The
    applet could support keys as long as the underlying Java Card
    implementation does, but the current personalisation APDU
    interface limits the amount of data when loading up private keys,
    which in effect limits the key size to 1024 bits. This limitation
    will hopefully be lifted in future versions of the applet.} The
  supported ciphers are the following:
\begin{itemize}
\item for signing (perform security operation command): RSA signature
  with PKCS 1.5 padding and SHA1 or SHA256 digests and RSA signature
  with PSS padding with SHA1 digest.  In all the cases the hashes have
  to provided ready to the card and in correct format, see APDU
  interface below, i.e.\ the card does not do the hashing, as
  stipulated by the ISO7816-8~\cite{ISO8} specification.\footnote{%
    Technical remark: because the hashes are provided ready to the
    card, the Java Card API \texttt{Signature} API could not be
    used. In turn, the PSS padding is done by a manually implemented
    method in the applet and involves the use of
    \texttt{MessageDigest.ALG\_SHA}.} The result of the signing
  operation is the RSA signature.  The corresponding Object
  Identifiers for the supported algorithms are:
  \begin{itemize}
    \item OID RSA SHA1 = 1.2.840.113549.1.1.5
    \item OID RSA SHA256 = 1.2.840.113549.1.1.11
    \item OID RSA PSS = 1.2.840.113549.1.1.10
    \item OID SHA1 = 1.3.14.3.2.26
    \item OID SHA256 = 2.16.840.1.101.3.4.2.1
  \end{itemize}
  The Java Card API involved is:
  \begin{itemize}
    \item \texttt{Cipher.ALG\_RSA\_PKCS1}
      \texttt{Cipher.ALG\_RSA\_NOPAD}
  \end{itemize}
\item for decryption (perform security operation command): RSA cipher
  with PKCS 1.5 padding. The input for this operation is a valid
  (i.e.\ properly formatted and with matching length) encrypted RSA
  block. The result is the decrypted plain text. The corresponding
  Object Identifier for this operation is:
  \begin{itemize}
    \item OID RSA = 1.2.840.113549.1.1.1
  \end{itemize}
  The Java Card API involved is:
  \begin{itemize}
    \item \texttt{Cipher.ALG\_RSA\_PKCS1}
  \end{itemize}
\item for authentication (internal authenticate command): RSA cipher
  with PKCS 1.5\break padding. The input for this operation is an arbitrary
  plain text with in the limits of the supported key length. The
  output is the encrypted RSA block.  The corresponding Object
  Identifier for this operation is:
  \begin{itemize}
    \item OID RSA = 1.2.840.113549.1.1.1
  \end{itemize}
  The Java Card API involved is:
  \begin{itemize}
    \item \texttt{Cipher.ALG\_RSA\_PKCS1}
  \end{itemize}
\end{itemize}
\item The AID of the applet is chosen to be
  \texttt{A000000063504B43532D3135}. The applet does not support\slash
  provide any FCI information on applet\slash file selection. It is
  expected that all such information required for the proper
  functioning of the host side application is stored in the file
  system in the form of ISO7816-15 structures (EF.DIR, EF.CIAInfo,
  EF.OD, EF.CD, etc.).  The host library and application provides an
  example of initialisation of such structures and uploading them to
  the card. It is also suggested that the PKI is made default
  selectable on the card.
\end{itemize}

\section{APDU Interface}

Below the APDUs that the applet supports are briefly described.

\subsection{Initialisation}

After a fresh applet is loaded onto the card, the suggested
initialisation sequence is the following:
\begin{enumerate}
  \item Select the applet.
  \item Set the state of the applet to \textsf{initial}. Applet is in
    personalisation mode.
  \item Optionally, change the historical bytes of the card through
    the applet. For this the applet needs to be default selectable.
  \item Load up the file structure information. \textbf{Note:} this
    step does not create any files, only provides the intended file
    structure information.
  \item Load up the RSA private keys and their identifiers to the
    card. Three keys are expected in total: for signing, decryption,
    and authentication.
  \item Create the contents of all the ISO7816-15 structures intended
    to be on the card (including user certificates matching the keys
    and CA certificate), create the corresponding files in the applet
    and load up the file contents to the card.
  \item Upload the PUC to the card. The PUC is unchangeable.
  \item Set the state of the applet to \textsf{prepersonalised}.
  \item At this stage the personalisation should be finished by
    setting up the user PIN (PUC has to be provided).  This will put
    the card into the \textsf{personalised} state. From this point on
    the personalised state of the applet is not changeable.
\end{enumerate}

The APDUs needed for that are described in the following.
Personalisation APDUs are only available in the \textsf{initial}
state.  On error conditions the APDUs may return a variety of abnormal
termination status words. On success the response APDU is always
\texttt{9000} with no data; during the personalisation phase no status
words other than \texttt{9000} should be accepted.

\subsubsection{Select Applet}\label{sec:selectaid}

\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{A4} & \texttt{04} & \texttt{00} &
\texttt{0B} & AID=\texttt{A000000063504B43532D3135} & Absent \\
\hline
\end{tabular}
\end{flushleft}


\subsubsection{Put Data -- Set Applet State}

\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{DA} & \texttt{68} & State &
\texttt{00} & Absent & Absent \\
\hline
\end{tabular}
\end{flushleft}
The P2/state can be one of: \texttt{01} initial, \texttt{02}
prepersonalised. The state is set to personalised \texttt{03}
implicitly by the change reference data command when setting the user
PIN.

\subsubsection{Put Data -- Upload File System Structure Information}

\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{DA} & \texttt{69} & \texttt{00} &
Var & Data with file structure information & Absent \\
\hline
\end{tabular}
\end{flushleft}
The data field in this APDU contains the file structure information
according to the following format. It is a list of concatenated single
file information byte strings.  For DF file the following bytes should
be present:
\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|}
\hline
$b_1$ & $b_2$ & $b_3$ & $b_4$ & $b_5$ & $b_6\dots b_n$ \\
\hline
\texttt{80} ($-1$) & $\mathrm{FID}_{\mathrm{MSB}}$ & $\mathrm{FID}_{\mathrm{LSB}}$ & 
Parent index & \#children & children indexes\\
\hline
\end{tabular}
\end{flushleft}
The indexes (parent and children) are relative to the beginning of the
whole data field. For EF files the format is this:
\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|}
\hline
$b_1$ & $b_2$ & $b_3$ & $b_4$ & $b_5$  \\
\hline
\texttt{00} & $\mathrm{FID}_{\mathrm{MSB}}$ & $\mathrm{FID}_{\mathrm{LSB}}$ & 
Parent index & SFI \\
\hline
\end{tabular}
\end{flushleft}
The SFI byte should be \texttt{00} if no SFI is provided for the file.
And example of a valid file system structure information is this
(quote from Java code):
\begin{verbatim}
byte[] fileStructure = {
  -1,            // DF
  0x3F, 0x00,    // FID, MF
  -1,            // no parent
  2, 7, 12,      // two children at indexes 7 and 12
  0,             // EF
  0x2F, 0x00,    // FID, EF.DIR
  0, 0x1E,       // parent at index 0, SFI is 1E
  -1,            // DF
  0x50, 0x15,    // FID, DF.CIA
  0,             // parent at index 0
  9, 26, 31, 36, 41, 46, 51, 56, 61, 66,
                 // 9 children
  0,             // EF
  0x50, 0x32,    // FID, EF.CIAInfo
  12, 0x12,      // parent at index 12, SFI is 12
  0, 0x50, 0x31, 12, 0x11, // EF.OD
  0, 0x42, 0x00, 12, 0x00, // EF.AOD
  0, 0x40, 0x00, 12, 0x00, // EF.PrKD
  0, 0x41, 0x00, 12, 0x00, // EF.CD
  0, 0x41, 0x01, 12, 0x00, // EF.CACert
  0, 0x41, 0x02, 12, 0x00, // EF.UserCert1
  0, 0x41, 0x03, 12, 0x00, // EF.UserCert2
  0, 0x41, 0x04, 12, 0x00, // EF.UserCert3
};
\end{verbatim}

\subsubsection{Put Data -- Set Historical Bytes}

\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{DA} & \texttt{67} & \texttt{00} &
Var & Historical bytes & Absent \\
\hline
\end{tabular}
\end{flushleft}

\subsubsection{Put Data -- Setup Private Key Identifier}

\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{DA} & \texttt{61}/\texttt{62}/\texttt{63} & \texttt{00} &
Var & Key identifier bytes & Absent \\
\hline
\end{tabular}
\end{flushleft}
The P1 byte indicates correspondingly: authentication key, signing
key, decryption key. The data field contains a key identifier, maximum
length 16 bytes.

\subsubsection{Put Data -- Upload Private Key}

\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{DA} & \texttt{64}/\texttt{65}/\texttt{66} & Key part &
Var & Corr.\ key part raw data & Absent \\
\hline
\end{tabular}
\end{flushleft}
The P1 byte indicates correspondingly: authentication key, signing
key, decryption key.  The P2 byte indicates the part of the raw data
of the key in CRT format in the data field, in the following
way:\footnote{It is my intention to simplify this key loading to use
  just the modulus and private exponent, i.e.\ the non-CRT form of the
  key. This will not be part of the current version of the applet
  though.}
\begin{flushleft}
\begin{tabular}{|l|l|}
\hline
Tag & Key part \\
\hline
\texttt{81} & public modulus\\
\texttt{82} & public exponent\\
\texttt{83} & prime number $P$\\
\texttt{84} & prime number $Q$\\
\texttt{85} & prime exponent $P$\\
\texttt{86} & prime exponent $Q$\\
\texttt{87} & CRT coefficient\\
\hline
\end{tabular}
\end{flushleft}

\subsubsection{Create File}

\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{E0} & \texttt{00} & \texttt{00} &
\texttt{05} & File FID, File length, PIN  & Absent \\
\hline
\end{tabular}
\end{flushleft}
The data field has the following format:
\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|}
\hline
$\mathit{FID}_{\mathrm{MSB}}$ & $\mathit{FID}_{\mathrm{LSB}}$ &
$\mathit{Len}_{\mathrm{MSB}}$ & $\mathit{Len}_{\mathrm{LSB}}$ &
PIN byte \\
\hline
\end{tabular}
\end{flushleft}
where PIN byte indicates whether the file should be PIN protected when
reading, \texttt{01} means PIN protect, \texttt{00} means free read.

\subsubsection{Select File}\label{sec:select}

Absolute, by file identifier \textit{FID} (if none provided, selects
MF):
\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{A4} & \texttt{00} & \texttt{00} & \texttt{02}/\texttt{00} &
$\mathit{FID}_{\mathrm{MSB}}$ $\mathit{FID}_{\mathrm{LSB}}$ or Absent & \texttt{00} \\
\hline
\end{tabular}
\end{flushleft}
DF (P1 is \texttt{01}) or EF (P1 is \texttt{02}) under the currently selected file, by file identifier \textit{FID}:
\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{A4} & \texttt{01}/\texttt{02} & \texttt{00} & \texttt{02} &
$\mathit{FID}_{\mathrm{MSB}}$ $\mathit{FID}_{\mathrm{LSB}}$ & \texttt{00} \\
\hline
\end{tabular}
\end{flushleft}
Parent of the current DF:
\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{A4} & \texttt{03} & \texttt{00} & \texttt{00} &
Absent & \texttt{00} \\
\hline
\end{tabular}
\end{flushleft}
By path from the current DF (P1 is \texttt{09}) or MF (P1 is
\texttt{08}):
\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{A4} & \texttt{08}/\texttt{09} & \texttt{00} & Var &
The file path & \texttt{00} \\
\hline
\end{tabular}
\end{flushleft}

\subsubsection{Write Binary}

\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{D0} &
$\mathit{Off}_{\mathrm{MSB}}$ or SFI & $\mathit{Off}_{\mathrm{LSB}}$ &
Var & Data to be written to the file & Absent \\
\hline
\end{tabular}
\end{flushleft}
The offset should to be less or equal than \texttt{7FFF} (most
significant bit set to 0). If the most significant bit of P1 is set to
1, then the remaining bits of P1 indicate SFI, in which case P2
indicates the offset.


\subsubsection{Change Reference Data -- Set PUC}

\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{24} & \texttt{01} & \texttt{00} &
\texttt{10} & PUC data & Absent \\
\hline
\end{tabular}
\end{flushleft}
This form of this command is only available when the applet is in the
\textsf{initial} state. The data field should contain the ASCII bytes
of the PUC number (i.e.\ from the range \texttt{30}\dots\texttt{39}).


\subsubsection{Change Reference Data -- Set PIN}\label{sec:pin}

\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{24} & \texttt{00} & \texttt{00} &
\texttt{14}\dots\texttt{24} & PUC data, PIN data & Absent \\
\hline
\end{tabular}
\end{flushleft}
The data field should contain the ASCII bytes of the PUC and PIN
numbers (i.e.\ from the range \texttt{30}\dots\texttt{39}).  The PUC
length should be always 16, the PIN length should be between 4 and 20.
If not already in the \textsf{personalised} state the applet state is
changed to \textsf{personalised} by this command.

\subsection{Communication after Personalisation}

After the personalisation the following set of ISO7816 commands is
available for regular applet operation. In the following, if not
stated otherwise the commands return a status word (\texttt{9000} on
success), and no response data.

\subsubsection{Select Applet}

See~\ref{sec:selectaid}.

\subsubsection{Select File}

See~\ref{sec:select}.

\subsubsection{Change Reference Data -- Change PIN}

See~\ref{sec:pin}.

\subsubsection{Read Binary}

\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{B0} &
$\mathit{Off}_{\mathrm{MSB}}$ or SFI & $\mathit{Off}_{\mathrm{LSB}}$ &
\texttt{00} & Absent & Response length \\
\hline
\end{tabular}
\end{flushleft}
The offset should to be less or equal than \texttt{7FFF} (most
significant bit set to 0). If the most significant bit of P1 is set to
1, then the remaining bits of P1 indicate SFI, in which case P2
indicates the offset.

This command will return the number of bytes specified by the Le field
in the command APDU. If there is no Le bytes available in the file,
the available bytes will be returned instead and the status word of
the response APDU will be \texttt{6282} (End of File).

\subsubsection{Verify PIN}

\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{20} & \texttt{00} & \texttt{00} &
\texttt{04}\dots\texttt{14} & PIN data & Absent \\
\hline
\end{tabular}
\end{flushleft}
The data field should contain the ASCII bytes of the PIN number
(i.e.\ from the range \texttt{30}\dots\texttt{39}).

\subsubsection{Manage Security Environment}

This command is used to prepare the upcoming crypto operation (Perform
Security Operation or Internal Authenticate). Just prior to the actual
crypto operation PIN verification is required (every time).
\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{22} & \texttt{41} & \texttt{A4}/\texttt{B6}/\texttt{B8} &
Var & Key Identifier DO, Algorithm Identifier DO & Absent \\
\hline
\end{tabular}
\end{flushleft}
The meaning of P2 is the following:
\begin{flushleft}
\begin{tabular}{|l|l|}
\hline
P2 & Security Operation \\
\hline
\texttt{A4} & Authentication \\
\texttt{B6} & Signing \\
\texttt{B8} & Decryption \\
\hline
\end{tabular}
\end{flushleft}
The data objects are formed as follows. The key identifier object is a
simple TLV structure with tag \texttt{84}. The data field of this TLV
structure contains the key identifier bytes.  The algorithm identifier
object is a simple TLV structure with tag \texttt{80}. The data field
of this TLV structure contains one byte corresponding to the crypto
algorithm. Usually, the information about possible values for this
byte and corresponding algorithms is included in the EF.CIAInfo file
on the card.

\subsubsection{Internal Authenticate}

This command requires prior successful Manage Security Environment and
Verify PIN commands.
\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{88} & \texttt{00} & \texttt{00} &
Var & The raw data to be signed\slash encrypted & Absent \\
\hline
\end{tabular}
\end{flushleft}
The maximum length of the command data is limited by the APDU length
(255), the length of the RSA key to be used and the padding algorithm
(key size in bytes minus 11). On success, the data in the response
APDU is the RSA cipher block.

\subsubsection{Perform Security Operation -- Sign}

This command requires prior successful Manage Security Environment and
Verify PIN commands.
\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{2A} & \texttt{9E} & \texttt{9A} &
Var & Hash to be signed & Absent \\
\hline
\end{tabular}
\end{flushleft}
Depending on the signing algorithm established by the prior MSE
command, the data field is one of the two:
\begin{itemize}
  \item A matching (SHA1 or SHA256) ASN.1 wrapped hash data object for
    RSA PKCS 1.5 signing algorithm.
  \item Raw SHA1 hash data for RSA PSS signing algorithm.
\end{itemize}
On success, the data in the response APDU is the RSA signature block.

\subsubsection{Perform Security Operation -- Decipher}

This command requires prior successful Manage Security Environment and
Verify PIN commands.  This is the only command that supports chaining
(CLA=\texttt{10}).
\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00}/\texttt{10} & \texttt{2A} & \texttt{80} & \texttt{82}/\texttt{84}/\texttt{86} &
Var & The cipher block to be decrypted & Absent \\
\hline
\end{tabular}
\end{flushleft}
The value of P2 can be any of the three, it does not influence the
operation of this command.  The data field is the (part of) cipher
block to be decrypted.  If the command is not last in the sequence the
response APDU is simply the status word.  After the last command
(CLA=\texttt{00}), on success, the data in the response APDU is the
decrypted plain text without padding.

\subsubsection{Get Challenge}

\begin{flushleft}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
CLA & INS & P1 & P2 & Lc & Data & Le \\
\hline
\texttt{00} & \texttt{84} & \texttt{00} & \texttt{00} &
\texttt{00} & Absent & Challenge length \\
\hline
\end{tabular}
\end{flushleft}
This command returns a random challenge from the card of the required
length. Currently this command has no interaction whatsoever with any
other crypto commands in the applet and is implemented for possible
future use.

\begin{thebibliography}{1}
\bibitem{ISO4} ISO/IEC.  \newblock Identification cards -- Integrated circuit cards --
  Part 4: Organization, security and commands for interchange.  \newblock Technical report,
  2005.  \newblock ISO 7816-4.

\bibitem{ISO8} ISO/IEC.  \newblock Identification cards -- Integrated circuit cards --
  Part 8: Commands for security operations.  \newblock Technical report,
  2004.  \newblock ISO 7816-8.

\bibitem{ISO15} ISO/IEC.  \newblock Identification cards -- Integrated circuit cards --
  Part 15: Cryptographic information application. \newblock Technical report,
  2004.  \newblock ISO 7816-15.


\end{thebibliography}

\end{document}
